Interfacing logic programming and Coq

### 2020-10-27 papers/types2020/sec3.tex
fig with dep kerned in sequent form added

\subsection{Horn logic in coq-elpi}

@@

\begin{footnotesize}
  \begin{figure}
  \newcommand{\XXi}{}
  \newcommand{\bc}[2]{#1 >\!> #2}
\[
\infer{\XXi\vdash t = t}
{%\eqExpert{\XXi}
}
\qquad
\infer{\XXi\vdash \true}
{%\trueExpert{\XXi}
}
\qquad
\infer{\XXi\vdash G_1\vee G_2}
{\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
}
\qquad
\infer{\XXi\vdash \exists x. G}
{\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
}
\qquad 
\infer{\XXi\vdash G_1\wedge G_2}
{\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{\XXi\vdash A}
      {\bc D A \quad D \in \Pscr}
\]
\vskip -18pt
\[
  \infer{\bc A A}{}{}\qquad
  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
\]
\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{A proof system for Horn logic and its encoding in coq-elpi .}
\label{fig:interp}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
%\caption{A dependent kernel for Coq terms.}
%\label{fig:kernel}
\end{figure}
\end{footnotesize}

\subsection{Focused proof systems with certificates}
\label{ssec:focused}

%
The notion of \emph{focused proof systems}~\cite{andreoli92jlc,liang09tcs}
generalizes this view of proof construction by identifying the
following two phases. % of proof construction.
\begin{enumerate}
\item The \emph{negative}%
% \footnote{The terminology of negative and positive phases is a bit
%   unfortunate: historically, these terms do not refer to positive or negative
%   subformula occurrences but rather to certain semantic models used in
%   the study of linear logic~\cite{girard87tcs}.}
%
  phase corresponds to  goal-reduction: in this phase,
  inference rules that involve \emph{don't-care-non\-de\-ter\-min\-ism} are
  applied.
%
  As a result, there is no need to consider backtracking over choices
  made in building this phase.

\item The \emph{positive} phase corresponding to  backchaining: in this phase, inference rules that involve
  \emph{don't-know-non\-de\-ter\-min\-ism} are applied: here,
  inference rules need to be supplied with information in order to
  ensure that a completed proof can be built.
%
  That information can be items such as which term is needed to
  instantiate a universally quantified formula and which disjunct of a
  disjunctive goal formula should be proved.
\end{enumerate}
%
Thus, when building a proof tree (in the sequent calculus) from the
conclusion to its leaves, the negative phase corresponds to a simple
computation that needs no external information, while the positive
phase may need such external information to be supplied.
%
In the literature, it is common to refer to a repository of such
external information as either an \emph{oracle} or a \emph{proof
certificate}.

\todo{Talk about the LJT thing: grammar for formulae and proof terms\cite{Herbelin94,LengrandDM06}?}

%\subsection{Certificate checking with expert predicates}
%\subsection{Adding certificates and experts to a proof system}
\label{ssec:fpc}

@@

  \newcommand{\bc}[2]{#1 >\!> #2}
\[
\infer{\XXi\vdash t = t}
{%\eqExpert{\XXi}
}
\qquad
\infer{\XXi\vdash \true}
{%\trueExpert{\XXi}
}
\qquad
\infer{\XXi\vdash G_1\vee G_2}
{\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
}
\qquad
\infer{\XXi\vdash \exists x. G}
{\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
}
\qquad 
\infer{\XXi\vdash G_1\wedge G_2}
{\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{\XXi\vdash A}
      {\bc D A \quad D \in \Pscr}
\]
\vskip -18pt
\[
  \infer{\bc A A}{}{}\qquad
  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
\]
\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

Figure~\ref{fig:augmented} contains a simple proof system for Horn
clause provability in which each inference rule is augmented with an
additional premise involving an \emph{expert predicate}, a
certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
case of $\vee$ and $\exists$).
%
%
% 
Although this proof system is a focused proof system, the
richness of focusing is not apparent in this simplified 
setting: thus, we drop the adjective ``focused'' from this point forward.

Figure~\ref{fig:kernel} contains the \lP implementation of the
inference rules in Figure~\ref{fig:augmented}: here the infix
turnstile $\vdash$ symbol is replaced by the \lst<i{check} predicate.
%
Notice that it is easy to show that no matter how the expert predicates
are defined, if the goal \lsti{check Cert B} is provable in \lP then
\lsti{B} must be a sound consequence of the program clauses stored in
\dots
% the \lsti{prog} predicate (which provides a natural implementation of
% the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
% AM: not there
% in Figure~\ref{fig:kernel}).

As we mentioned in the introduction, our notion of proof certificates
 is taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:kernel}:
that is, the exact set of constructors for the \lsti{cert} type as
well as the exact specification of the six expert predicates listed in
that figure.
%
Figure~\ref{fig:resources} displays two such FPCs,
both of which can be used to describe proofs where we bound
the number of occurrences of unfoldings in a proof.
%
For example, the first FPC
provides the experts for treating certificates that are constructed
using the \lsti{qheight} constructor.
%
As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} (for the
encoding \lsti{B} of a goal formula) is provable in \lP using the
clauses in Figures~\ref{fig:kernel} and~\ref{fig:resources} if and
only if the height of that proof is 5 or less.
%
Similarly, the second FPC uses the constructor \lsti{qsize} (with two
integers) and can be used to bound the total number of instances of
unfoldings in a proof.
%
In particular, the query \mbox{\lsti{sigma H\ (check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
is provable if and only if the total number of unfoldings of that
proof is 5 or less.

@@

Various  additional examples and experiments using the pairing of FPCs
can be found in~\cite{Pair}. Using similar techniques, it is possible to define FPCs that target
specific types for special treatment: for example, when generating
integers, only (user-defined) small integers could be produced.

>>>>

Stuff in red is from old paper
\subsection{Horn logic in coq-elpi}

@@

\begin{footnotesize}
  \begin{figure}
  \newcommand{\XXi}{}
  \newcommand{\bc}[2]{#1 \Downarrow #2}
\[
\infer{\XXi\vdash t = t}
{%\eqExpert{\XXi}
}
\qquad
\infer{\XXi\vdash \true}
{%\trueExpert{\XXi}
}
\qquad
\infer{\XXi\vdash G_1\vee G_2}
{\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
}
\qquad
\infer{\XXi\vdash \exists x. G}
{\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
}
\qquad 
\infer{\XXi\vdash G_1\wedge G_2}
{\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{\XXi\vdash A}
      {\bc D A \quad D \in \Pscr}
\]
\vskip -18pt
\[
  \infer{\bc A A}{}{}\qquad
  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
\]
\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{A proof system for Horn logic and its encoding in coq-elpi .}
\label{fig:interp}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
%\caption{A dependent kernel for Coq terms.}
%\label{fig:kernel}
\end{figure}
\end{footnotesize}

\subsection{Focused proof systems with certificates}
\label{ssec:focused}
\todo{Adapt the discussion}
\todo{Talk about the LJT thing: grammar for formulae and proof terms\cite{Herbelin94,LengrandDM06}?}

%
{\color{red}
The notion of \emph{focused proof systems}~\cite{andreoli92jlc,liang09tcs}
generalizes this view of proof construction by identifying the
following two phases. % of proof construction.
\begin{enumerate}
\item The \emph{negative}%
% \footnote{The terminology of negative and positive phases is a bit
%   unfortunate: historically, these terms do not refer to positive or negative
%   subformula occurrences but rather to certain semantic models used in
%   the study of linear logic~\cite{girard87tcs}.}
%
  phase corresponds to  goal-reduction: in this phase,
  inference rules that involve \emph{don't-care-non\-de\-ter\-min\-ism} are
  applied.
%
  As a result, there is no need to consider backtracking over choices
  made in building this phase.

\item The \emph{positive} phase corresponding to  backchaining: in this phase, inference rules that involve
  \emph{don't-know-non\-de\-ter\-min\-ism} are applied: here,
  inference rules need to be supplied with information in order to
  ensure that a completed proof can be built.
%
  That information can be items such as which term is needed to
  instantiate a universally quantified formula and which disjunct of a
  disjunctive goal formula should be proved.
\end{enumerate}
%
Thus, when building a proof tree (in the sequent calculus) from the
conclusion to its leaves, the negative phase corresponds to a simple
computation that needs no external information, while the positive
phase may need such external information to be supplied.
%
In the literature, it is common to refer to a repository of such
external information as either an \emph{oracle} or a \emph{proof
certificate}.
}

%\subsection{Certificate checking with expert predicates}
%\subsection{Adding certificates and experts to a proof system}
%\label{ssec:fpc}

@@

  \newcommand{\bc}[5]{#1;#2\vdash #3 : #4 \Downarrow #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
\infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
{\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{\XXi;\Gamma\vdash t : A}
      {\bc{\XXi}{\Gamma} t D A \quad D \in \Pscr\cup\Gamma}
\]
\vskip -18pt
\[
  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{\Gamma}{(t\ l)} {\forall x:B.\, D} A}
  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

Figure~\ref{fig:augmented} contains a simple proof system \dots  in which each inference rule is augmented with an
additional premise involving an \emph{expert predicate}, a
certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
case of $\vee$ and $\exists$).
%
%
%
\todo{Explain relations with other term assignments. Note that we do not use explicit substitutions (we are cut free). Explain the choice of not having certificates represent proof terms. Explain that the unfold clause handles defined atoms and defined connectives (conjunction, disjunction, existential and also equality: this may be less efficient and proof-theoretically induce pointless phase changes, but it is arguably very elegant e simplifies proof terms that are inherited by Coq }

Figure~\ref{fig:kernel} contains the elpi implementation of the
inference rules in Figure~\ref{fig:augmented}: here the infix
turnstile $\vdash$ symbol is replaced by the \lsti{check} predicate.
%
% Notice that it is easy to show that no matter how the expert predicates
% are defined, if the goal \lsti{check Cert B} is provable in \lP then
% \lsti{B} must be a sound consequence of the program clauses stored in
% \dots
% the \lsti{prog} predicate (which provides a natural implementation of
% the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
% AM: not there
% in Figure~\ref{fig:kernel}).

\paragraph{Certificates}

\todo{Shorten it}
\todo{Show the dprolog tactic? have an example? could  be
\lsti{Goal  exists R, insert 2 [0;1] R.} and we would move the code for insert here}
{\color{red}
As we mentioned in the introduction, our notion of proof certificates
 is taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:kernel}:
that is, the exact set of constructors for the \lsti{cert} type as
well as the exact specification of the six expert predicates listed in
that figure.
%
Figure~\ref{fig:resources} displays two such FPCs,
both of which can be used to describe proofs where we bound
the number of occurrences of unfoldings in a proof.
%
For example, the first FPC
provides the experts for treating certificates that are constructed
using the \lsti{qheight} constructor.
%
As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} (for the
encoding \lsti{B} of a goal formula) is provable in \lP using the
clauses in Figures~\ref{fig:kernel} and~\ref{fig:resources} if and
only if the height of that proof is 5 or less.
%
Similarly, the second FPC uses the constructor \lsti{qsize} (with two
integers) and can be used to bound the total number of instances of
unfoldings in a proof.
%
In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
is provable if and only if the total number of unfoldings of that
proof is 5 or less.

@@

Various  additional examples and experiments using the pairing of FPCs
can be found in~\cite{Pair}. Using similar techniques, it is possible to define FPCs that target
specific types for special treatment: for example, when generating
integers, only (user-defined) small integers could be produced.
}
